A Bayesian Perspective \\\large{Extracorporeal CPR for Refractory Out-of-Hospital Cardiac Arrest}

### 2023-02-12 manuscript/manuscript.qmd
convergence added

---
title: A Bayesian Perspective 
subtitle: Extracorporeal CPR for Refractory Out-of-Hospital Cardiac Arrest
author:
  - name: James M Brophy MD PhD
    email: james.brophy@mcgill.ca
    affiliations: 
        - id: McGill University 
          name: McGill University Health Center
          department: Centre for Health Outcomes Research (CORE) 
          address: 5252 Boul. de Maisonneuve West Room 2B.37
          city: Montreal
          state: QC
          postal-code: H4A 3S5
    attributes:
        corresponding: true
    note: JMB is a research scholar supported by Les Fonds de Recherche Québec Santé
abstract: |
  A recent randomized clinical trial, INCEPTION, reported in patients with refractory out-of-hospital cardiac arrest, extracorporeal CPR (eCPR) and conventional CPR (cCPR) had similar effects on survival with a favorable neurologic outcome. The current study examines if a Bayesian perspective provides  additional insights. Depending on the prior selected, the Bayesian approach for the INCEPTION intention-to treat (ITT) analysis shows an equivalence probability between 13.4 - 16.8% (defined as 1 / 1.1 < odds ratio (OR) < 1.1). The probability of clinical superiority with eCPR ranges from 65.7 - 77.0 % (defined as OR > 1.1). A similar analyses using INCEPTION per protocol (PP) data shows an equivalence probability between 4.7 - 20.2% with reduced probabilities of clinical superiority not exceeding 25%. It is concluded that a Bayesian perspective allows considerable additional insights into the trial analysis and interpretation. The non-negligible probabilities of increased survival, or even harm, with their considerable residual uncertainties, suggests that additional studies are required before concluding that eCPR and cCPR have similar survival effects.
keywords: 
  - extracorporeal CPR
  - Bayesian statistics
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: To be determined
      formatting: preprint
      model: 3p
      cite-style: super
editor: 
  markdown: 
    wrap: 72
---

@@

Posterior distributions were estimated using the Hamiltonian Monte Carlo,a form of 
Markov Chain Monte Carlo simulations in which the gradient of the log posterior is 
used to efficiently sample the posterior space. This was implemented in `Stan`[@stan] using the front end `rstanarm` package[@rstanarm] by fitting a logistic regression model with a single treatment parameter. All analyses were executed using `R`[@R]  within the integrated development environmentof RStudio[@Rstudio].  The statistical code can be found on Github
(https://github.com/brophyj/eCPR).

@@

# - INCEPTION alone default vague priors
set.seed(1234)
fit1 <- stan_glm(prop_success ~ Tx, data = df[c(1:2),], family = binomial(link="logit"), weights = total, refresh=0) #stan_glm with combined prior
fit1_sum <- summary(fit1, digits=3)
conf_95 <- c(exp(fit1_sum[2]),exp(fit1_sum[2]+c(-1,1)*1.96*fit1_sum[10]))
tt <- fit1 %>%
    spread_draws(`(Intercept)`, `TxeCPR`)
fig_prob <- sum(tt$TxeCPR>0)/nrow(tt)
# summary(fit1, digits=3)
# print(c(exp(.2785),exp(.2785+c(-1,1)*1.96*.4537)))
# get_variables(fit1)
# prior_summary(fit1)
# help('prior_summary.stanreg') 
# P > 1.1
fig_prob_1.1 <- sum(tt$TxeCPR > 0.0953)/nrow(tt) # OR >1.1 (log (1.1) = 0.0953)
# [1] 0.657

@@

Although this  Bayesian analysis presents the strengths noted above, it also had limitations. First, we did not have access to either
individual data from any of the trials limiting the possibility of examining specific subgroups. Also for the per protocol analysis, this data was only available for survival and not the primary outcome that included survival with intact neurological status. The threshold choices for superiority and equivalence may be seen as arbitrary, although the analyses can be easily repeated for different choices. Given the existence of only 3 trials  differences between the baseline cCPR rates could not be reliably modeled. We choose not to perform a traditional Bayesian meta-analysis with
a semi-informative prior for the between-study variation  because our
goal was to illustrate how Bayesian principles could be informative
when applied to the interpretation of new trial data.

The overall message from this study is that standard statistical analyses leading to a conclusion of "similar survival effects" is overly simplified and potentially inaccurate. This Bayesian analysis demonstrates clearly that while any definitive conclusions regarding the superiority of eCPR are lacking, the possibility of a clinically meaningful benefit, or even less likely the possibility of clinically meaningful harm, has not been reasonably excluded and continued research is necessary to clarifying the remaining uncertainties. The final dilemma confronting clinicians is that even a Bayesian analysis of a randomized trial provides only probability estimates of an average treatment effect and not the elusive individual treatment effect.

>>>>

---
title: A Bayesian Perspective 
subtitle: Extracorporeal CPR for Refractory Out-of-Hospital Cardiac Arrest
author:
  - name: James M Brophy MD PhD
    email: james.brophy@mcgill.ca
    affiliations: 
        - id: McGill University 
          name: McGill University Health Center
          department: Centre for Health Outcomes Research (CORE) 
          address: 5252 Boul. de Maisonneuve West Room 2B.37
          city: Montreal
          state: QC
          postal-code: H4A 3S5
    attributes:
        corresponding: true
    note: JMB is a research scholar supported by Les Fonds de Recherche Québec Santé
abstract: |
  A recent randomized clinical trial, INCEPTION, reported in patients with refractory out-of-hospital cardiac arrest, extracorporeal CPR (eCPR) and conventional CPR (cCPR) had similar effects on survival with a favorable neurologic outcome. The current study examines if a Bayesian perspective provides  additional quantative insights. Depending on the prior selected, the Bayesian approach for the INCEPTION intention-to treat (ITT) analysis shows an equivalence probability between 13.4 - 16.8% (defined as 1 / 1.1 < odds ratio (OR) < 1.1). The probability of clinical superiority with eCPR ranges from 65.7 - 77.0 % (defined as OR > 1.1). A similar analyses using INCEPTION per protocol (PP) data shows an equivalence probability between 4.7 - 20.2% with reduced probabilities of clinical superiority not exceeding 25%. It is concluded that a Bayesian perspective allows considerable additional quantative insights into the trial analysis and interpretation. The non-negligible probabilities of increased survival, or even harm, with their considerable residual uncertainties, suggests that additional studies are required before concluding that eCPR and cCPR have similar average survival effects.
keywords: 
  - extracorporeal CPR
  - Bayesian statistics
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: To be determined
      formatting: preprint
      model: 3p
      cite-style: super
editor: 
  markdown: 
    wrap: 72
---

@@

Posterior distributions were estimated using the Hamiltonian Monte Carlo,a form of 
Markov Chain Monte Carlo simulations in which the gradient of the log posterior is 
used to efficiently sample the posterior space. This was implemented in `Stan`[@stan] using the front end `rstanarm` package[@rstanarm] by fitting a logistic regression model with a single treatment parameter. All analyses were executed using `R`[@R]  within the integrated development environmentof RStudio[@Rstudio].  Model convergence was assessed by examination of the Monte Carlo Standard Error < 10% of the posterior standard deviation, $n_{eff}$ an estimate of the effective number of independent draws from the posterior distribution of the estimand > 10% maximum and $\hat{R}$  a measures the ratio of the average variance of samples within each chain to the variance of the pooled samples across chains < 1.1. The statistical code can be found on Github
(https://github.com/brophyj/eCPR).

@@

# - INCEPTION alone default vague priors
set.seed(1234)
fit1 <- stan_glm(prop_success ~ Tx, data = df[c(1:2),], family = binomial(link="logit"), weights = total, refresh=0) #stan_glm with combined prior
fit1_sum <- summary(fit1, digits=3) # check convergence
conf_95 <- c(exp(fit1_sum[2]),exp(fit1_sum[2]+c(-1,1)*1.96*fit1_sum[10]))
tt <- fit1 %>%
    spread_draws(`(Intercept)`, `TxeCPR`)
fig_prob <- sum(tt$TxeCPR>0)/nrow(tt)
# summary(fit1, digits=3)
# print(c(exp(.2785),exp(.2785+c(-1,1)*1.96*.4537)))
# get_variables(fit1)
# prior_summary(fit1)
# help('prior_summary.stanreg') 
# plot(fit1)
# P > 1.1
fig_prob_1.1 <- sum(tt$TxeCPR > 0.0953)/nrow(tt) # OR >1.1 (log (1.1) = 0.0953)
# [1] 0.657

@@

Although this  Bayesian analysis presents the strengths noted above, it also had limitations. First, we did not have access to either
individual data from any of the trials limiting the possibility of examining specific subgroups. Also for the per protocol analysis, this data was only available for survival and not the primary outcome that included survival with intact neurological status. The threshold choices for superiority and equivalence may be seen as arbitrary, although the analyses can be easily repeated for different choices. Given the existence of only 3 trials  estimating the very divergent baseline cCPR rates could not be reliably modeled. We choose not to perform a traditional Bayesian meta-analysis with
a semi-informative prior for the between-study variation  because our
goal was to illustrate how Bayesian principles could be informative
when applied to the analysis and interpretation of new trial data, both in the situation of no and considerable previous knowledge.

The overall message from this study is that standard statistical analyses leading to a conclusion of "similar survival effects" may be overly simplified and potentially inaccurate. This Bayesian analysis demonstrates clearly that while any definitive conclusions regarding the superiority of eCPR are lacking, the possibility of a clinically meaningful benefit, or even less likely the possibility of clinically meaningful harm, has not been reasonably excluded and continued research is necessary to clarifying the remaining uncertainties. The final dilemma confronting clinicians is that even a Bayesian analysis of a randomized trial provides only probability estimates of an average treatment effect and not the elusive individual treatment effect.

### 2023-02-12 manuscript/manuscript.tex
convergence added

\begin{abstract}
A recent randomized clinical trial, INCEPTION, reported in patients with
refractory out-of-hospital cardiac arrest, extracorporeal CPR (eCPR) and
conventional CPR (cCPR) had similar effects on survival with a favorable
neurologic outcome. The current study examines if a Bayesian perspective
provides additional insights. Depending on the prior selected, the
Bayesian approach for the INCEPTION intention-to treat (ITT) analysis
shows an equivalence probability between 13.4 - 16.8\% (defined as 1 /
1.1 \textless{} odds ratio (OR) \textless{} 1.1). The probability of
clinical superiority with eCPR ranges from 65.7 - 77.0 \% (defined as OR
\textgreater{} 1.1). A similar analyses using INCEPTION per protocol
(PP) data shows an equivalence probability between 4.7 - 20.2\% with
reduced probabilities of clinical superiority not exceeding 25\%. It is
concluded that a Bayesian perspective allows considerable additional
insights into the trial analysis and interpretation. The non-negligible
probabilities of increased survival, or even harm, with their
considerable residual uncertainties, suggests that additional studies
are required before concluding that eCPR and cCPR have similar survival
effects.
\end{abstract}

\begin{keyword}
    extracorporeal CPR \sep 
    Bayesian statistics
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

Posterior distributions were estimated using the Hamiltonian Monte
Carlo,a form of Markov Chain Monte Carlo simulations in which the
gradient of the log posterior is used to efficiently sample the
posterior space. This was implemented in \texttt{Stan}\citep{stan} using
the front end \texttt{rstanarm} package\citep{rstanarm} by fitting a
logistic regression model with a single treatment parameter. All
analyses were executed using \texttt{R}\citep{R} within the integrated
development environmentof RStudio\citep{Rstudio}. The statistical code
can be found on Github (https://github.com/brophyj/eCPR).

@@

Although this Bayesian analysis presents the strengths noted above, it
also had limitations. First, we did not have access to either individual
data from any of the trials limiting the possibility of examining
specific subgroups. Also for the per protocol analysis, this data was
only available for survival and not the primary outcome that included
survival with intact neurological status. The threshold choices for
superiority and equivalence may be seen as arbitrary, although the
analyses can be easily repeated for different choices. Given the
existence of only 3 trials differences between the baseline cCPR rates
could not be reliably modeled. We choose not to perform a traditional
Bayesian meta-analysis with a semi-informative prior for the
between-study variation because our goal was to illustrate how Bayesian
principles could be informative when applied to the interpretation of
new trial data.

The overall message from this study is that standard statistical
analyses leading to a conclusion of ``similar survival effects'' is
overly simplified and potentially inaccurate. This Bayesian analysis
demonstrates clearly that while any definitive conclusions regarding the
superiority of eCPR are lacking, the possibility of a clinically
meaningful benefit, or even less likely the possibility of clinically
meaningful harm, has not been reasonably excluded and continued research
is necessary to clarifying the remaining uncertainties. The final
dilemma confronting clinicians is that even a Bayesian analysis of a
randomized trial provides only probability estimates of an average
treatment effect and not the elusive individual treatment effect.

>>>>

\begin{abstract}
A recent randomized clinical trial, INCEPTION, reported in patients with
refractory out-of-hospital cardiac arrest, extracorporeal CPR (eCPR) and
conventional CPR (cCPR) had similar effects on survival with a favorable
neurologic outcome. The current study examines if a Bayesian perspective
provides additional quantative insights. Depending on the prior
selected, the Bayesian approach for the INCEPTION intention-to treat
(ITT) analysis shows an equivalence probability between 13.4 - 16.8\%
(defined as 1 / 1.1 \textless{} odds ratio (OR) \textless{} 1.1). The
probability of clinical superiority with eCPR ranges from 65.7 - 77.0 \%
(defined as OR \textgreater{} 1.1). A similar analyses using INCEPTION
per protocol (PP) data shows an equivalence probability between 4.7 -
20.2\% with reduced probabilities of clinical superiority not exceeding
25\%. It is concluded that a Bayesian perspective allows considerable
additional quantative insights into the trial analysis and
interpretation. The non-negligible probabilities of increased survival,
or even harm, with their considerable residual uncertainties, suggests
that additional studies are required before concluding that eCPR and
cCPR have similar average survival effects.
\end{abstract}

\begin{keyword}
    extracorporeal CPR \sep 
    Bayesian statistics
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, frame hidden, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

Posterior distributions were estimated using the Hamiltonian Monte
Carlo,a form of Markov Chain Monte Carlo simulations in which the
gradient of the log posterior is used to efficiently sample the
posterior space. This was implemented in \texttt{Stan}\citep{stan} using
the front end \texttt{rstanarm} package\citep{rstanarm} by fitting a
logistic regression model with a single treatment parameter. All
analyses were executed using \texttt{R}\citep{R} within the integrated
development environmentof RStudio\citep{Rstudio}. Model convergence was
assessed by examination of the Monte Carlo Standard Error \textless{}
10\% of the posterior standard deviation, \(n_{eff}\) an estimate of the
effective number of independent draws from the posterior distribution of
the estimand \textgreater{} 10\% maximum and \(\hat{R}\) a measures the
ratio of the average variance of samples within each chain to the
variance of the pooled samples across chains \textless{} 1.1. The
statistical code can be found on Github
(https://github.com/brophyj/eCPR).

@@

Although this Bayesian analysis presents the strengths noted above, it
also had limitations. First, we did not have access to either individual
data from any of the trials limiting the possibility of examining
specific subgroups. Also for the per protocol analysis, this data was
only available for survival and not the primary outcome that included
survival with intact neurological status. The threshold choices for
superiority and equivalence may be seen as arbitrary, although the
analyses can be easily repeated for different choices. Given the
existence of only 3 trials estimating the very divergent baseline cCPR
rates could not be reliably modeled. We choose not to perform a
traditional Bayesian meta-analysis with a semi-informative prior for the
between-study variation because our goal was to illustrate how Bayesian
principles could be informative when applied to the analysis and
interpretation of new trial data, both in the situation of no and
considerable previous knowledge.

The overall message from this study is that standard statistical
analyses leading to a conclusion of ``similar survival effects'' may be
overly simplified and potentially inaccurate. This Bayesian analysis
demonstrates clearly that while any definitive conclusions regarding the
superiority of eCPR are lacking, the possibility of a clinically
meaningful benefit, or even less likely the possibility of clinically
meaningful harm, has not been reasonably excluded and continued research
is necessary to clarifying the remaining uncertainties. The final
dilemma confronting clinicians is that even a Bayesian analysis of a
randomized trial provides only probability estimates of an average
treatment effect and not the elusive individual treatment effect.
